\rs{Kink and fluting instabilities in a dynamically twisted flux tube}{Flute and kink instabilities in a dynamically twisted flux tube with anisotropic plasma viscosity}

### 2021-09-24 paper.tex
Add Rado's changes

\documentclass[12pt]{article}
\usepackage[T1]{fontenc} % modern encoding
\usepackage[utf8]{inputenc}
\usepackage{libertine}
\usepackage[libertine]{newtxmath}
\usepackage[scaled=0.95]{inconsolata}
\usepackage[margin=0.5in]{geometry}

@@

%% --- user packages ---

\usepackage{csquotes}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{grffile}
\usepackage{overpic}
\usepackage{float}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{easy-todo}
\usepackage{listings}

@@

% For editing
\usepackage{color}
\newcommand{\rr}[1]{\textcolor{red}{#1}}
\newcommand{\bb}[1]{\textcolor{blue}{#1}}
\newcommand{\rs}[1]{\textcolor{blue}{#1}}
%\newcommand{\rr}[1]{#1}
%\newcommand{\bb}[1]{#1}
%\newcommand{\rs}[1]{#1}

@@

Kink and fluting instabilities in a dynamically twisted flux tube

@@

@@

The helical kink instability is a form of ideal magnetohydrodynamic (MHD) instability which occurs in highly twisted magnetic flux tubes such as those making up much of the solar corona~\cite{realeCoronalLoopsObservations2014} and has been well studied in the coronal context~\cite{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003c,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020a}. Given its energetic nonlinear development, it is considered a potential mechanism for heating the solar corona through the theory of nanoflares~\cite{klimchukSolvingCoronalHeating2006a,browningMechanismsSolarCoronal1991} and a key mechanism in the production of solar flares~\cite{hoodKinkInstabilitySolar1979}. Previous work by the same authors has investigated a twisted magnetic flux tube already linearly unstable to the helical kink instability, investigating particularly the effect of anisotropic viscosity on the nonlinear dynamics~\cite{quinnEffectAnisotropicViscosity2020a}. In~\cite{quinnEffectAnisotropicViscosity2020a} and most other investigations of the kink instability, a perturbation is applied to an already significantly twisted flux tube. An alternative way to excite the instability (and the way employed here) is to start with an initially straight field and apply twisting motions at the boundaries to form a twisted flux tube which eventually becomes unstable. This kind of dynamic excitation of the kink instability is useful in that it represents more closely the real development of a magnetic flux tube and associated instabilities in the solar corona. The dynamic twisting of the flux tube reveals an additional instability, the fluting instability, which has been theorised (for example, in~\cite{priestMagnetohydrodynamicsSuna}) but not yet observed in coronal simulations or observations, to our knowledge.

The fluting instability arises in magnetised plasmas where the plasma pressure gradient is directed in the same direction as the field line curvature, that is the pressure and magnetic tension forces compete. This is similar to the competition between pressure and gravitational forces which gives rise to the Rayleigh-Taylor instability (RTI). In MHD, the RTI is a typical example of an ideal interchange instability, where magnetic field lines are minimally bent and are, instead, exchanged during the evolution of the instability. The ideal fluting instability is another example of an ideal interchange instability but confined to a cylindrical geometry, the name ``fluting instability'' referring to its likeness to a fluted column. In a twisted flux tube like a simple, unbraided coronal loop, the magnetic curvature is always directed towards the axis so the tube may be unstable to fluting when the pressure decreases outwards from a high-pressure core. Such a pressure distribution is generated in the flux tubes studied here as a result of the driving. The appearance of the fluting instability is illustrated by, for example, the pressure contours in figure~\ref{fig:kink_field_line_plots}, where the perturbations follow the pitch of the twisted field.

In other solar contexts, interchange instabilities can be found in the form of ballooning modes in arcades~\cite{hoodBallooningInstabilitiesSolar1986}, as the instability which forms tubes of specific size in the photosphere~\cite{bunteInterchangeInstabilitySolar1993} and in the buoyancy of flux tubes~\cite{schuesslerInterchangeInstabilitySmall1984}. However, the fluting instability specifically is more commonly studied in fusion contexts~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}. In fusion, the focus is generally on understanding how a particular plasma device may be stabilised to the instability in particular geometries such as that of the mirror machine~\cite{jungwirthTheoryFluteInstability1965} or in toroidal geometries such as the tokomak~\cite{shafranovFluteInstabilityCurrentcarrying1968}. The resistive fluting instability (also known as the resistive interchange instability) can be excited even when the ideal fluting instability is stabilised. As a result, this has been given significantly more attention~\cite{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}. While this body of research is useful and applicable in solar contexts, it is mostly limited to the study of the stability and linear development of the fluting instability, the nonlinear development being of secondary importance to the functioning of fusion devices. More detailed investigations of its nonlinear development is required to understand its importance in the context of coronal dynamics and coronal heating. The development of the fluting instability and its interaction with the simultaneously growing kink instability is the focus of this work and the experiments described here represent an initial exploration into the nonlinear fluting instability in the solar corona. Of particular interest is the effect of anisotropic viscosity, which is found to strongly affect the growth of the fluting instability.

It is well known that viscosity in magnetised plasmas (such as that which makes up the solar corona) is anisotropic and strongly dependent on the strength and direction of the local magnetic field~\cite{hollwegViscosityChewGoldbergerLowEquations1986a,hollwegViscosityMagnetizedPlasma1985,Braginskii1965}. The development of the switching model~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017} and implementation in Lare3d has allowed the investigation of the effect of anisotropic viscosity on solar phenomena. We find the stability of Lare3d without (the more commonly employed) isotropic or shock viscosity to be sufficient for at least the study of kink instabilities in coronal loops~\cite{quinnEffectAnisotropicViscosity2020a}. The interest in anisotropic viscosity stems from the unanswered question of which heating mechanism (viscous or Ohmic) is dominant in the solar corona~\cite{klimchukSolvingCoronalHeating2006a}, an important facet of solving the coronal heating problem. Using scaling laws, it has been suggested that viscous heating (generated through anisotropic viscosity) can dwarf that of Ohmic heating~\cite{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}. However, due to computational and observational limitations, this cannot be directly tested, and so the influence of other factors such as small scale instabilities and turbulence is relatively unknown~\cite{klimchukSolvingCoronalHeating2006a}. In addition to directly heating the plasma, viscosity plays a part in the damping of instabilities and waves~\cite{rudermanSlowSurfaceWave2000a}. It is this effect we are most interested in here, and it shall be found that the use of anisotropic viscosity permits the growth of the fluting instability, which is otherwise strongly damped by isotropic viscosity.

\subsection{The fluting instability}

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers in the $\theta$ and $z$ directions, respectively. The helical kink instability occurs for perturbations where $m=1, k\ne0$ and is the only instability of this form which is a body instability, that is it moves the entire body of the flux tube. Perturbations where $m>1$ are termed fluting or interchange instabilities.

When the magnetic field is sheared, as in a twisted magnetic flux tube, an interchange instability (such as the fluting instability) is confined to a surface where the peaks and troughs follow the shear of the field. That is, the instability is confined to the surface where the perturbation wavevector $(0, m/r, k)$ is perpendicular to the direction of the field which, in an axisymmetric, twisted flux tube, is located at a radius $r$ given by
\begin{equation}
  \label{eq:resonant_surface}
\frac{m}{r} B_{\theta}(r) + kB_z(r) \approx 0.
\end{equation}

The stability of an infinite cylindrical flux tube to perturbations of the form~\eqref{eq:kink_perturbation} is given by the classical Suydam's criterion
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L B_{\theta}$ is the safety factor for a flux tube of length $L$ and a dash denotes differentiation with respect to $r$~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies to both fluting and kink instabilities although many additional effects such as line-tying are not incorporated into the corresponding linear analysis. The effect of line-tying on the kink instability can be found in~\cite{hoodKinkInstabilitySolar1979}. Where~\eqref{eq:suydams_criterion} is not satisfied, the flux tube may be unstable to perturbations of the form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations remain local to resonant surfaces given by~\eqref{eq:resonant_surface}.

When Suydam's criterion is satisfied and the flux tube is linearly stable, it may still be unstable to non-local perturbations, where the shear and pressure are small enough that interchange perturbations do not need to satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity generally reduces the stabilising effect of the shear, permitting growth of a resistive interchange mode, albeit at a slower rate than that of the ideal instability~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be found that the ideal linear analysis of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for understanding the fluting instabilities investigated here since the associated flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

While Suydam's condition gives an indication of the stability of a flux tube to a given perturbation, the linear growth rate of the ideal fluting instability $\gamma$ can be found via a stability analysis analogous to that of the Rayleigh-Taylor instability (see~\cite{goldstonIntroductionPlasmaPhysics2020}) and is given by
\begin{equation}
  \label{eq:fluting_growth_rate}
\gamma^2 = \frac{2|\nabla p|}{\rho R_c},
\end{equation}
where $R_c$ is the radius of curvature of the magnetic field. This equation only applies when the pressure gradient and radius of curvature vector are in the same direction, that is the plasma is constrained by a concave magnetic field such that the pressure forces and magnetic tension forces are in competition. In a cylindrical, twisted flux tube, the field is always concave towards the central axis of the tube, so any inwardly directed pressure gradient is potentially unstable to fluting.

Throughout this chapter, the twisted flux tube generated by the drivers has a pressure profile which is approximately axisymmetric, and independent of $z$ away from the boundaries at $z=\pm2$, and has a negative gradient, hence $|\nabla p|$ may be written as $-d p/ dr$. Similarly, away from the boundaries, the magnetic field has a negligible $r$ component and little dependence on $\theta$ and $z$, allowing the field to be approximated as $\vec{B} = (0, B_{\theta}(r), B_z(r))^{\text{T}}$, in cylindrical coordinates $(r, \theta, z)$. For a twisted field of this form, the radius of curvature is given by 
\begin{equation}
  \label{eq:radius_of_curvature}
  R_c = \frac{1}{|(\vec{b}\cdot\nabla) \vec{b}|} = \frac{r}{b_{\theta}^2},
\end{equation}
where $\vec{b} = \vec{B}/|\vec{B}|$ is the unit vector in the direction of the magnetic field and $b_{\theta}$ is the component of $\vec{b}$ in the azimuthal direction. These approximations allows the growth rate to be written as
\begin{equation}
  \label{eq:fluting_growth_rate2}
\gamma_{ideal}^2 = \frac{-2p'}{\rho R_c}.
\end{equation}
This growth rate remains applicable while the flux tube remains relatively axisymmetric, that is while the kink instability remains in its linear phase.

In contrast to the precise form of the equilibrium state and perturbation studied in~\cite{quinnEffectAnisotropicViscosity2020a}, this is an experiment where a system is driven and instabilities occur organically as a result of noise providing a random perturbation. As a result of the driving, the flux tube is not in static equilibrium. Consequently, the stability criterion and linear growth rate presented previously are useful only as a guide and for approximate comparison.

\section{Numerical setup}

Lare3d numerically solves the MHD equations for the density $\rho$, plasma velocity $\vec{u}$, pressure $p$, magnetic field $\vec{B}$ and internal energy $\varepsilon$,
\begin{gather}
%\begin{equation}
\label{eq:mhda}
\frac{D\rho}{Dt} = - \rho \vec{\nabla} \cdot \vec{u},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\vec{u}}{Dt} = -\vec{\nabla} p + \vec{\jmath} \times \vec{B} + \vec{\nabla} \cdot \ten{\sigma},\\
%\end{equation}
%\begin{equation}
\frac{D\vec{B}}{Dt} = (\vec{B} \cdot \vec{\nabla})\vec{u} - (\vec{\nabla} \cdot \vec{u})\vec{B} + \eta \nabla^2 \vec{B},\\
%\end{equation}
%\begin{equation}
\rho\frac{D\varepsilon}{Dt} = -p \vec{\nabla} \cdot \vec{u} + {Q}_{\nu} + {Q}_{\eta},%\\
\label{eq:energy}
%\end{equation}
\end{gather}
where $\eta$ is the non-dimensionalised resistivity, $\jmath = \nabla \times \vec{B}$ is the current density and $\ten{\sigma}$ is the viscous stress tensor (described below). The terms ${Q}_{\nu} = \ten{\sigma} : \vec{\nabla}\vec{u}$ and ${Q}_{\eta} = \eta | \vec{\jmath} |^2$ are viscous heating and Ohmic heating, respectively. The system is closed by the inclusion of the equation of state for an ideal gas
\begin{equation}
\varepsilon = \frac{p}{\rho(\gamma - 1)},
\end{equation}
where the specific heat ratio is given by $\gamma = 5/3$.

The isotropic (or Newtonian) viscosity stress tensor is given by
\begin{equation}
  \label{eq:isotropic_viscous_tensor}
  \ten{\sigma} = \nu \ten{W},
\end{equation}
where $\nu$ is the viscous transport parameter, generally referred to as the viscosity,
\begin{equation}
  \label{eq:rate_of_strain}
  \ten{W} = \nabla\vec{u} + (\nabla\vec{u})^T - \tfrac{2}{3}(\nabla \cdot \vec{u})\ten{I},
\end{equation}
is the rate of strain tensor, and $\ten{I}$ is the  $3\times 3$ identity matrix. We use an identical model of anisotropic viscosity as that used in previous work~\cite{quinnEffectAnisotropicViscosity2020a}, the switching model, first presented in~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}. While this model is intended to revert to an isotropic model in the vicinity of magnetic null points, we effectively only make use of the anisotropic component, since there are no null points present in the simulations at any point. To this end, we set the switching parameter (which controls the degree of viscous anisotropy) to $a_0 = 150$ in Lare3d's control, a value sufficiently large that the anisotropic viscous stress tensor is given by
\begin{equation}
  \label{eq:switching_model}
\ten{\sigma}_{\text{swi}} = \nu \left[\frac{3}{2}(\ten{W}\vec{b}\cdot\vec{b}) \left( \vec{b} \otimes \vec{b} - \frac{1}{3}\ten{I} \right)\right],
\end{equation}
where $\vec{b}$ is the unit vector in the direction of the magnetic field. This tensor is identical to the strong field approximation of Braginskii~\cite{braginskiiTransportProcessesPlasma1965}. Full use of the switching model can be found in~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017,quinnEffectAnisotropicViscosity2020a}\todo{ref KHI paper}.

The non-dimensionalisation scheme and values align with the scheme used by Lare3d~\cite{arberStaggeredGridLagrangian2001} and are identical to those used in~\cite{quinnEffectAnisotropicViscosity2020a}. The magnetic field is prescribed as initially straight and uniform,
\begin{equation}
\vec{B} = (0, 0, 1)^{\text{T}},
\end{equation}
in a cube of dimension $[-2,2]^3$. The velocity is set everywhere to $\vec{u} = \vec{0}$, the density to $\rho = 1$, the internal energy to $\varepsilon = 8.67\times 10^{-4}$ (corresponding to a temperature of $10^6$ K). \todo{what's the plasma beta?} At the boundaries, the magnetic field, density, and energy are fixed to their initial values and the velocity to zero except at the boundaries where the twisting driver, described below, is prescribed. The spatial derivatives of these variables are also set to zero at the boundaries. The resolution is $512$ grid points per dimension, comparable to the highest resolution kink instability studies of~\cite{hoodCoronalHeatingMagnetic2009} or medium resolution studies of~\cite{barefordShockHeatingNumerical2015}.

\begin{figure}[t]
  \centering
  \begin{subfigure}{.49\textwidth}
  \centering
  \includegraphics[width=1.0\linewidth]{u_r.pdf}
  \caption{Radial dependence of driver}
  \label{fig:kink_radial_driver}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
  \centering
  \includegraphics[width=1.0\linewidth]{u_t.pdf}
  \caption{Acceleration of driver}
  \label{fig:kink_driver_accel}
  \end{subfigure}

  \mycaption{Radial velocity profile $u_r(r)$ and acceleration profile $u_t(t)$ of the driver for parameters $u_0 = 0.15$, $r_d = 5$ and $t_r = 2$.}{}
  \label{fig:kink_driver}
\end{figure}

>>>>

\documentclass[12pt]{article}
%\usepackage[T1]{fontenc} % modern encoding
%\usepackage[utf8]{inputenc}
%\usepackage{libertine}
%\usepackage[libertine]{newtxmath}
%\usepackage[scaled=0.95]{inconsolata}
\usepackage[margin=0.5in]{geometry}

@@

%% --- user packages ---
\usepackage{amssymb}
\usepackage{csquotes}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{grffile}
\usepackage{overpic}
\usepackage{float}
\usepackage{enumitem}
\usepackage{amsfonts}
\usepackage{easy-todo}
\usepackage{listings}

@@

% For editing
\usepackage{color}
\newcommand{\rr}[1]{\textcolor{red}{#1}}
\newcommand{\bb}[1]{\textcolor{blue}{#1}}
\newcommand{\rs}[2]{\textcolor{red}{#2}}
%\newcommand{\rr}[1]{#1}
%\newcommand{\bb}[1]{#1}
%\newcommand{\rs}[1]{#1}

@@

\title{\rs{Kink and fluting instabilities in a dynamically twisted
    flux tube}{Flute and kink instabilities in a dynamically twisted
    flux tube with anisotropic plasma viscosity}}
\author{\rs{}{James J.~Quinn$^{1,\ast}$, Radostin D.~Simitev$^2$}}
\date{\small
  $^1$ Research Software Development Group, University College London, Gower Street, London WC1E 6BT, UK.\\
  e-mail:
  \href{mailto:jamiejquinn@jamiejquinn.com}{jamiejquinn@jamiejquinn.com}. $^\ast$
  Corresponding author\\
  $^2$ School of Mathematics and Statistics, University of Glasgow,
  Glasgow G12 8QQ,
  UK. \\ \href{https://orcid.org/0000-0002-2207-5789}{orcid.org/0000-0002-2207-5789}\\
\vspace{5mm}
  Version: \today}
\maketitle

@@

\begin{abstract}
  \rs{}{NEEDS ABSTRACT}
\end{abstract}

@@

The helical kink instability is a form of ideal magnetohydrodynamic
(MHD) instability which occurs in highly twisted magnetic flux tubes
such as those making up much of the solar
corona~\cite{realeCoronalLoopsObservations2014} and has been well
studied in the coronal
context~\cite{hoodKinkInstabilitySolar1979,hoodCoronalHeatingMagnetic2009,browningSolarCoronalHeating2003c,barefordShockHeatingNumerical2015,quinnEffectAnisotropicViscosity2020a}. Given
its energetic nonlinear development, it is considered a potential
mechanism for heating the solar corona through the theory of
nanoflares~\cite{klimchukSolvingCoronalHeating2006a,browningMechanismsSolarCoronal1991}
and a key mechanism in the production of solar
flares~\cite{hoodKinkInstabilitySolar1979}. Previous work by the same
authors has investigated a twisted magnetic flux tube already linearly
unstable to the helical kink instability, \rs{investigating particularly}{focussing specifically on}
the effect of anisotropic viscosity on the nonlinear
dynamics~\cite{quinnEffectAnisotropicViscosity2020a}. In~\cite{quinnEffectAnisotropicViscosity2020a}
and most other investigations of the kink instability
\rs{}{e.g. \cite{hoodCoronalHeatingMagnetic2009}}, a perturbation 
is applied to an already significantly twisted flux tube. An
alternative way to excite the instability (and the way employed here)
is to start with an initially straight field and apply twisting
motions at the boundaries to form a twisted flux tube which eventually
becomes unstable. This kind of dynamic excitation of the kink
instability is useful in that it represents more closely \rs{the real
development of a magnetic flux tube and associated}{actual evolution
  of magnetic flux tubes and the associated} instabilities in 
the solar corona. \rs{The}{In our simulations, the}
dynamic twisting of the flux tube reveals an additional instability,
the \rs{fluting}{flute} instability, which has been theorised (for example,
in~\cite{priestMagnetohydrodynamicsSuna}) but not yet observed in
coronal simulations or observations, to our knowledge. \rs{}{RADO: NOT
  YET OBSERVED? 
  WHAT ABOUT \cite{terradasEffectMagneticTwist2018,rudermanNonlinearGenerationFluting2017a}??!!]}

The \rs{fluting}{flute} instability arises in magnetised plasmas where
the plasma pressure gradient is \rs{directed}{oriented} in the same direction as the
field line curvature, that is the pressure and magnetic tension forces
compete. This is similar to the competition between pressure and
gravitational forces which gives rise to the Rayleigh-Taylor
instability (RTI). In \rs{MHD}{magnetohydrodynamic terminology}, the RTI is a typical example of an ideal
interchange instability, where magnetic field lines are minimally bent
and are, instead, exchanged during the evolution of the
instability. The ideal \rs{fluting}{flute} instability is another example of an
ideal interchange instability but confined to a cylindrical geometry,
the name \rs{``fluting instability''}{``flute instability''} referring to its likeness to a fluted
column. In a twisted flux tube like a simple, unbraided coronal loop,
the magnetic curvature is always directed towards the axis so the tube
may be unstable to fluting when the pressure decreases outwards from a
high-pressure core. Such a pressure distribution is generated in the
flux tubes studied here as a result of the driving. The appearance of
the \rs{fluting}{flute} instability is illustrated by, for example, the pressure
contours in figure~\ref{fig:kink_field_line_plots}, where the
perturbations follow the pitch of the twisted field.

In other solar contexts, interchange instabilities can be found in the
form of ballooning modes in
arcades~\cite{hoodBallooningInstabilitiesSolar1986}, as the
instability which forms tubes of specific size in the
photosphere~\cite{bunteInterchangeInstabilitySolar1993} and in the
buoyancy of flux
tubes~\cite{schuesslerInterchangeInstabilitySmall1984}. However, the
\rs{fluting}{flute} instability specifically is more commonly studied in fusion
contexts~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998,zhengAdvancedTokamakStability2015,wessonHydromagneticStabilityTokamaks1978}. In
fusion, the focus is generally on understanding how a particular
plasma device may be stabilised to the instability in particular
geometries such as that of the mirror
machine~\cite{jungwirthTheoryFluteInstability1965} or in toroidal
geometries such as the
tokomak~\cite{shafranovFluteInstabilityCurrentcarrying1968}. The
resistive \rs{fluting}{flute} instability (also known as the resistive interchange
instability) can be excited even when the ideal \rs{fluting}{flute} instability is
stabilised. As a result, this has been given significantly more
attention~\cite{johnsonResistiveInterchangesNegativeV1967,correa-restrepoResistiveBallooningModes1983}. While
this body of research is useful and applicable in solar contexts, it
is mostly limited to the study of the stability and linear development
of the \rs{fluting}{flute} instability, the nonlinear development being of
secondary importance to the functioning of fusion devices. More
detailed investigations of its nonlinear development is required to
understand its importance in the context of coronal dynamics and
coronal heating. The development of the \rs{fluting}{flute} instability and its
interaction with the simultaneously growing kink instability is the
\rs{}{main} focus of this work and the experiments described here represent an
initial exploration into the nonlinear \rs{fluting}{flute} instability in the
solar corona. \rs{Of particular interest is the effect of anisotropic
viscosity, which is found to strongly affect the growth of the \rs{fluting}{flute}
instability.}{}

\rs{}{In addition to our main goal, of particular interest here is
also the effect of anisotropic plasma viscosity, which in the following is found to
strongly influence the growth of the \rs{fluting}{flute} instability.}
It is well known that viscosity in magnetised plasmas (such as
\rs{that which makes up}{those which make up} the solar corona) is
anisotropic and strongly dependent on the strength and direction of
the local magnetic
field~\cite{hollwegViscosityChewGoldbergerLowEquations1986a,hollwegViscosityMagnetizedPlasma1985,braginskiiTransportProcessesPlasma1965}.
\rs{The development of the switching
model~\cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017}
and implementation in Lare3d has allowed the investigation of
the effect of anisotropic viscosity on solar phenomena.
We find the stability of Lare3d without (the more commonly employed)
isotropic or shock viscosity to be sufficient for at least the study
of kink instabilities in coronal loops~\cite{quinnEffectAnisotropicViscosity2020a}.
}{To take this into account, in \cite{mactaggartBraginskiiMagnetohydrodynamicsArbitrary2017},
we developed a phenomenological model of anisotropic viscosity that
captures the main physics of viscosity in the 
solar corona as outlined in the analysis of
\cite{braginskiiTransportProcessesPlasma1965}, namely parallel
viscosity in regions of strong field  strength and isotropic viscosity
in regions of very weak or zero field strength. For brevity, we will
refer to this model of viscosity as ``the switching model''. In
\cite{quinnEffectAnisotropicViscosity2020a,Quinn2021} we implemented
the switching model as a module for the widely-used general MHD code
Lare3d \cite{arberStaggeredGridLagrangian2001}, and demonstrated
significant effects of anisotropic viscosity on the 
development of the nonlinear MHD kink instability and the
Kelvin-Helmholtz instability.} \rs{The}{More generally, the} interest in
anisotropic viscosity stems from the \rs{unanswered}{open} question of
which heating mechanism (viscous or Ohmic) is dominant in the solar
corona~\cite{klimchukSolvingCoronalHeating2006a}, an important facet
of solving the coronal heating problem. Using scaling laws, it has
been suggested that viscous heating (generated through anisotropic
viscosity) can dwarf that of Ohmic
heating~\cite{craigAnisotropicViscousDissipation2009a,litvinenkoViscousEnergyDissipation2005}.
However,
due to computational and observational limitations, this cannot be
directly tested, and so the influence of other factors such as small
scale instabilities and turbulence is relatively
unknown~\cite{klimchukSolvingCoronalHeating2006a}. In addition to
directly heating the plasma, viscosity plays a part in the damping of
instabilities and waves~\cite{rudermanSlowSurfaceWave2000a}. It is
this effect we are most interested in here, and it shall be \rs{found}{reported} that
the use of anisotropic viscosity permits the growth of the
\rs{fluting}{flute} instability, which is otherwise strongly damped by
isotropic viscosity.

\rs{}{ADD A BRIEF NEW ``CONTENTS PARAGRAPH'':
This paper is organized as follows. Section 1 ..., Section 2...}

\subsection{The \rs{fluting}{flute} instability}

In general, the stability of a cylindrical twisted magnetic flux tube is analysed using perturbations of the form
\begin{equation}
  \label{eq:kink_perturbation}
\xi(r, \theta, z) = \xi(r) e^{i(m\theta + kz)},
\end{equation}
where $m$ and $k$ are the wavenumbers \rs{in the $\theta$ and $z$
  directions, respectively}{in the azimuthal and axial directions,
  $\theta$ and $z$, respectively and $r$ is the radial coordinate in
  cylindrical polars}. 
The helical kink instability occurs for perturbations where $m=1, k\ne0$ and is the only instability of this form which is a body instability, that is it moves the entire body of the flux tube. Perturbations where $m>1$ are termed \rs{fluting}{flute} or interchange instabilities.

When the magnetic field is sheared, as in a twisted magnetic flux
tube, an interchange instability (such as the \rs{fluting}{flute}
instability) is confined to a surface where the peaks and troughs
follow the shear of the field. That is, the instability is confined to
the surface where the perturbation wavevector $(0, m/r, k)$ is
perpendicular to the direction of the field, \rs{}{known as the
  ``resonance surface''}. \rs{which in}{In} an axisymmetric twisted
flux tube \rs{}{the resonance surface} is located at a radius $r$
given by
\begin{equation}
  \label{eq:resonant_surface}
\frac{m}{r} B_{\theta}(r) + kB_z(r) \approx 0.
\end{equation}

The stability of an infinite cylindrical flux tube to perturbations of
the form~\eqref{eq:kink_perturbation} is given by the classical
\rs{}{Suydam's criterion \cite{Suydam1959}}
\begin{equation}
  \label{eq:suydams_criterion}
\frac{B_z^2 S^2}{4} + 2 r p' > 0,
\end{equation}
where $S = r q'/q$ is a measure of the shear, $q = 2\pi r B_z / L
B_{\theta}$ is the safety factor for a flux tube of length $L$ and a
\rs{dash}{prime} denotes differentiation with respect to
$r$~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998}. This applies
to both \rs{fluting}{flute} and kink instabilities although many
additional effects such as line-tying are not incorporated into the
corresponding linear analysis. The effect of line-tying on the kink
instability \rs{can be found}{is investigated}
in~\cite{hoodKinkInstabilitySolar1979}. Where~\eqref{eq:suydams_criterion}
is not satisfied, the flux tube may be unstable to perturbations of
the form~\eqref{eq:kink_perturbation}. When $m>1$, the perturbations
remain local to resonant surfaces given
by~\eqref{eq:resonant_surface}. 
\rs{}{When} Suydam's criterion is satisfied and the flux tube is linearly stable, it may still be unstable to non-local perturbations, where the shear and pressure are small enough that interchange perturbations do not need to satisfy~\eqref{eq:resonant_surface}. Additionally, the inclusion of resistivity generally reduces the stabilising effect of the shear, permitting growth of a resistive interchange mode, albeit at a slower rate than that of the ideal instability~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998}. It will be found that the ideal linear analysis of~\cite{mikhailovskiiInstabilitiesConfinedPlasma1998} is sufficient for understanding the \rs{fluting}{flute} instabilities investigated here since the associated flux tubes adequately fail the criterion~\eqref{eq:suydams_criterion}.

While Suydam's condition gives an indication of the stability of a flux tube to a given perturbation, the linear growth rate of the ideal \rs{fluting}{flute} instability $\gamma$ can be found via a stability analysis analogous to that of the Rayleigh-Taylor instability (see~\cite{goldstonIntroductionPlasmaPhysics2020}) and is given by
\begin{equation}
  \label{eq:fluting_growth_rate}
\gamma^2 = \frac{2|\nabla p|}{\rho R_c},
\end{equation}
where $R_c$ is the radius of curvature of the magnetic field. This equation only applies when the pressure gradient and radius of curvature vector are in the same direction, that is the plasma is constrained by a concave magnetic field such that the pressure forces and magnetic tension forces are in competition. In a cylindrical, twisted flux tube, the field is always concave towards the central axis of the tube, so any inwardly directed pressure gradient is potentially unstable to fluting.

Throughout this \rs{chapter}{paper}, the twisted flux tube generated by the drivers has a pressure profile which is approximately axisymmetric, and independent of $z$ away from the boundaries at $z=\pm2$, and has a negative gradient, hence $|\nabla p|$ may be written as $-d p/ dr$. Similarly, away from the boundaries, the magnetic field has a negligible $r$ component and little dependence on $\theta$ and $z$, allowing the field to be approximated as $\vec{B} = (0, B_{\theta}(r), B_z(r))^{\text{T}}$, in cylindrical coordinates $(r, \theta, z)$. For a twisted field of this form, the radius of curvature is given by 
\begin{equation}
  \label{eq:radius_of_curvature}
  R_c = \frac{1}{|(\vec{b}\cdot\nabla) \vec{b}|} = \frac{r}{b_{\theta}^2},
\end{equation}
where $\vec{b} = \vec{B}/|\vec{B}|$ is the unit vector in the direction of the magnetic field and $b_{\theta}$ is the component of $\vec{b}$ in the azimuthal direction. These approximations allows the growth rate to be written as
\begin{equation}
  \label{eq:fluting_growth_rate2}
\gamma_{\rs{{ideal}}{\text{ideal}}}^2 = \frac{-2p'}{\rho R_c}.
\end{equation}
This \rs{growth rate remains applicable}{approximation for the growth rate continues to hold} while the flux tube remains relatively axisymmetric, that is while the kink instability remains in its linear phase.

\rs{In contrast to the precise form of the equilibrium state and
  perturbation studied in~\cite{quinnEffectAnisotropicViscosity2020a},
  this is an experiment where a system is driven and instabilities
  occur organically as a result of noise providing a random
  perturbation. As a result of the driving, the flux tube is not in
  static equilibrium. Consequently, the stability criterion and linear
  growth rate presented previously are useful only as a guide and for
  approximate comparison.}{The stability criterion
  \eqref{eq:suydams_criterion} and the linear
  growth rate approximation \eqref{eq:fluting_growth_rate2} are useful
  only as a guide and for 
  approximate analysis of the numerical simulations presented in this
  work.  The precise form of the equilibrium state and the
  perturbations needed for the validity of
  \eqref{eq:suydams_criterion} and \eqref{eq:fluting_growth_rate2}
  were used in~\cite{quinnEffectAnisotropicViscosity2020a}.
  In contrast, in the experiments reported in the following the system
  is driven and instabilities occur spontaneously due to random
  perturbation. As a result of the driving, the flux tube is also not
  in static equilibrium initially. }

\section{\rs{Numerical setup}{Mathematical formulation and numerical setup}}
